---
title: "EI_comparison"
author: "Fabian Blombach"
date: "13/05/2019"
output: html_document
---
This script compares EI data between different growth phases
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
expon <- read.delim("data/data.expon.txt", header=T)
h2o2  <- read.delim("data/data.h2o2.txt", header=T)

require(ggplot2)
require(colorspace)
theme_set(theme_bw(base_size = 8))
```

make data frames combining EI data for the intersection of TUs in two data sets
```{r data frames with intersection of TUs between two conditions}
exph2o2<-merge(expon[,c(1,11,13,19,21,23:31,36:37,39:40, 49, 55, 63)], h2o2[,c(1,11,13,19,21,23:31,36:37,39:40)], by.x="operon_id", by.y="operon_id", sort=F)

print(nrow(exph2o2))
```

##EI value comparison with t-test exponential phase vs oxidative stress
```{r t.test_oxid}
#calculating p vales and padj for Welch's t-tests per gene
p.exp.rnap <- vector()
p.exp.spt45 <- vector()
p.exp.elf1 <- vector()
for (i in 1:nrow(exph2o2)){
  p <-t.test(c(exph2o2$EI_RNAP1.x[i], exph2o2$EI_RNAP2.x[i]), 
             c(exph2o2$EI_RNAP1.y[i], exph2o2$EI_RNAP2.y[i]))$p.val
  p.exp.rnap <- c(p.exp.rnap, p)
  q <-t.test(c(exph2o2$EI_spt45_r1.x[i], exph2o2$EI_spt45_r2.x[i]), 
             c(exph2o2$EI_spt45_r1.y[i], exph2o2$EI_spt45_r2.y[i]))$p.val
  p.exp.spt45 <- c(p.exp.spt45, q)
  r <-t.test(c(exph2o2$EI_elf1_r1.x[i], exph2o2$EI_elf1_r2.x[i]), 
             c(exph2o2$EI_elf1_r1.y[i], exph2o2$EI_elf1_r2.y[i]))$p.val
  p.exp.elf1 <- c(p.exp.elf1, r)
}
p.exp.rnap
sum(p.adjust(p.exp.rnap, method="BH") < 0.1)
exph2o2$padj.t.RNAP<- p.adjust(p.exp.rnap, method="BH") < 0.1

p.exp.spt45
sum(p.adjust(p.exp.spt45, method="BH") < 0.1)
exph2o2$padj.t.Spt45<- p.adjust(p.exp.spt45, method="BH") < 0.1

p.exp.elf1
sum(p.adjust(p.exp.elf1, method="BH") < 0.1)
exph2o2$padj.t.Elf1<- p.adjust(p.exp.elf1, method="BH") < 0.1
```


  
Scatter plots for TUs comparison to oxidative stress condition
```{r scatter expon vs h2o2}
#plotting
plot.data<-function(x,y,z) {
  range.xy <- max(c(exph2o2[[x]], exph2o2[[y]])) - min(c(exph2o2[[x]], exph2o2[[y]]))
  lm.cx<- ceiling(2*(max(c(exph2o2[[x]], exph2o2[[y]]))  + 0.05*range.xy))/2
  lm.fx<- floor(2*(min(c(exph2o2[[x]], exph2o2[[y]])) - 0.05*range.xy))/2
  lm.cy<-lm.cx
  lm.fy<-lm.fx
  plot <- ggplot(data=exph2o2, aes_string(x=x , y=y, colour=z))
  plot <- plot + geom_point(cex=0.2) + 
    xlab(x) + ylab(y) + 
    scale_y_continuous(limits=c(lm.fy, lm.cy), breaks=c(-10:10), expand = c(0, 0)) +
    scale_x_continuous(limits=c(lm.fx, lm.cx), breaks=c(-10:10), expand = c(0, 0)) + 
    theme(aspect.ratio=1,panel.grid.major = element_blank(), panel.grid.minor  = element_blank(), 
          legend.position = "none", axis.ticks = element_line(colour = "black", size = 0.5),
          panel.border  = element_rect(colour = "black", size = 0.5)) +
    ggtitle(paste0(x,".vs.", y)) +
    geom_abline(slope=1)
  plot
  #Saving plot as 45x45 mm pdf 
  ggplot2::ggsave(paste0("scatter.", x, "v", y, ".pdf"), plot = plot, device = "pdf", path = NULL,
                  scale = 1, width = 45, height = 45, units = c("mm"), limitsize = TRUE)
}


plot.data("EI_RNAPmean.x", "EI_RNAPmean.y", "padj.t.RNAP")
plot.data("EI_Spt45mean.x", "EI_Spt45mean.y", "padj.t.Spt45")
plot.data("EI_Elf1mean.x", "EI_Elf1mean.y", "padj.t.Elf1")
```


##Correlated changes between exponential growth phase and oxidative stress
```{r link_to_transcriptome}
correl.df <- data.frame(RNA.change = log(exph2o2$Expression.expon/exph2o2$Expression.h2o2,2),
                      TFB.change = log(sqrt(exph2o2$TFB1_foldR1.x * exph2o2$TFB1_foldR2.x),2) -
                        log(sqrt(exph2o2$TFB1_foldR1.y * exph2o2$TFB1_foldR2.y),2),
                      TFEb.change = log(sqrt(exph2o2$TFEb_foldR1.x * exph2o2$TFEb_foldR2.x),2) -
                        log(sqrt(exph2o2$TFEb_foldR1.y * exph2o2$TFEb_foldR2.y),2),
                      RNAP.EI.change = exph2o2$EI_RNAPmean.x - exph2o2$EI_RNAPmean.y,
                      Spt45.EI.change = exph2o2$EI_Spt45mean.x - exph2o2$EI_Spt45mean.y,
                      Elf1.EI.change = exph2o2$EI_Elf1mean.x - exph2o2$EI_Elf1mean.y,
                      CPSF1.load.change = log(exph2o2$cpsf1_r1_promoter.x/exph2o2$elf1_r1_promoter.x,2) +
                        log(exph2o2$cpsf1_r2_promoter.x/exph2o2$elf1_r2_promoter.x,2) -
                        log(exph2o2$cpsf1_r1_promoter.y/exph2o2$elf1_r1_promoter.y,2) -
                        log(exph2o2$cpsf1_r2_promoter.y/exph2o2$elf1_r2_promoter.y,2)
                      ) 
#filter out ribosomal promoter
correl.df<-correl.df[!is.na(correl.df$RNA.change),]


heatmap.df <- cor(correl.df, method = "spearman")
heatmap.df<-reshape2::melt(heatmap.df)

hm <- ggplot(data=heatmap.df, aes(x=Var1, y=Var2)) + geom_raster(aes(fill=value), interpolate = F) + 
   scale_fill_continuous_diverging("Purple-Brown", limits=c(-1,1), rev=T) + theme_void()
hm  
ggplot2::ggsave("heatmap.changes.expon.vs.h2o2.pdf", plot = hm, device = "pdf", path = NULL,
       scale = 1, width = 25, height = 25, units = c("mm"), limitsize = TRUE)

#pvalue calculation
pval.df<-matrix(nrow=7, ncol=7)
for(i in 1:7){
  for(j in 1:7){
    x<-cor.test(correl.df[,i], correl.df[,j], method="spearman")$p.value
    pval.df[i,j] <- x
    }
}

#multiple testing correction BH for unique tests
pval.v<-vector()
for(i in 1:6){
  x<-c(pval.df[i,(i+1):7])
  pval.v<-c(pval.v, x)
}

padj.v <- p.adjust(pval.v, "BH")
padj.v

#p 0.05 threshold
padj.v < 0.05
```


```{r seaborn plots}
require(reticulate)
use_python('~/miniconda3/bin/python3.7')

#importing required Python libraries/modules
sns <- import('seaborn')
plt <- import('matplotlib.pyplot')
pd <- import('pandas')

#recreate heatmap df
heatmap.df <- cor(correl.df, method = "spearman")

#building a heatmap using seaborn 
#r_to_py() converts R object into a python 
sns$heatmap(r_to_py(heatmap.df), fmt="g", cmap ="RdBu_r", center=0, vmin=-1, vmax=1)
#save the plot
plt$savefig("heatmap.seaborn.exponVsH2O2.pdf")
#display the plot
plt$show()
```


#Repeating analysis for two combinations of biological replicates for each condition
```{r link_to_transcriptome_repl1.h2o2}
correl.df.r1 <- data.frame(RNA.change = log(exph2o2$Expression.expon/exph2o2$Expression.h2o2,2),
                      TFB.change = log(exph2o2$TFB1_foldR1.x,2) -
                        log(exph2o2$TFB1_foldR1.y,2),
                      TFEb.change = log(exph2o2$TFEb_foldR1.x,2) -
                        log(exph2o2$TFEb_foldR1.y,2),
                      RNAP.EI.change = exph2o2$EI_RNAP1.x - exph2o2$EI_RNAP1.y,
                      Spt45.EI.change = exph2o2$EI_spt45_r1.x - exph2o2$EI_spt45_r1.y,
                      Elf1.EI.change = exph2o2$EI_elf1_r1.x - exph2o2$EI_elf1_r1.y,
                      CPSF1.load.change = log(exph2o2$cpsf1_r1_promoter.x/exph2o2$elf1_r1_promoter.x,2) -
                        log(exph2o2$cpsf1_r1_promoter.y/exph2o2$elf1_r1_promoter.y,2)
                      ) 

heatmap.df.r1 <- cor(correl.df.r1, method = "spearman")
heatmap.df.r1 <-reshape2::melt(heatmap.df.r1)

hm.r1 <- ggplot(data=heatmap.df.r1, aes(x=Var1, y=Var2)) + geom_raster(aes(fill=value), interpolate = F) + 
   scale_fill_continuous_diverging("Purple-Brown", limits=c(-1,1), rev=T) + theme_void()
hm.r1  
ggplot2::ggsave("heatmap.changes.expon.vs.h2o2.repl1.pdf", plot = hm.r1, device = "pdf", path = NULL,
       scale = 1, width = 25, height = 25, units = c("mm"), limitsize = TRUE)

#pvalue calculation

pval.df.r1<-matrix(nrow=7, ncol=7)
for(i in 1:7){
  for(j in 1:7){
    x<-cor.test(correl.df.r1[,i], correl.df.r1[,j], method="spearman")$p.value
    pval.df.r1[i,j] <- x
    }
}

#converting data frame to vector
pval.v.r1<-vector()
for(i in 1:6){
  x<-c(pval.df.r1[i,(i+1):7])
  pval.v.r1<-c(pval.v.r1, x)
}
```


```{r link_to_transcriptome_repl2.h2o2}
correl.df.r2 <- data.frame(RNA.change = log(exph2o2$Expression.expon/exph2o2$Expression.h2o2,2),
                      TFB.change = log(exph2o2$TFB1_foldR2.x,2) -
                        log(exph2o2$TFB1_foldR2.y,2),
                      TFEb.change = log(exph2o2$TFEb_foldR2.x,2) -
                        log(exph2o2$TFEb_foldR2.y,2),
                      RNAP.EI.change = exph2o2$EI_RNAP2.x - exph2o2$EI_RNAP2.y,
                      Spt45.EI.change = exph2o2$EI_spt45_r2.x - exph2o2$EI_spt45_r2.y,
                      Elf1.EI.change = exph2o2$EI_elf1_r2.x - exph2o2$EI_elf1_r2.y,
                      CPSF1.load.change = log(exph2o2$cpsf1_r2_promoter.x/exph2o2$elf1_r2_promoter.x,2) -
                        log(exph2o2$cpsf1_r2_promoter.y/exph2o2$elf1_r2_promoter.y,2)
                      ) 

heatmap.df.r2 <- cor(correl.df.r2, method = "spearman")
heatmap.df.r2 <-reshape2::melt(heatmap.df.r2)

hm.r2 <- ggplot(data=heatmap.df.r2, aes(x=Var1, y=Var2)) + geom_raster(aes(fill=value), interpolate = F) + 
   scale_fill_continuous_diverging("Purple-Brown", limits=c(-1,1), rev=T) + theme_void()
hm.r2  
ggplot2::ggsave("heatmap.changes.expon.vs.h2o2.repl2.pdf", plot = hm.r2, device = "pdf", path = NULL,
       scale = 1, width = 25, height = 25, units = c("mm"), limitsize = TRUE)

#pvalue calculation
pval.df.r2<-matrix(nrow=7, ncol=7)
for(i in 1:7){
  for(j in 1:7){
    x<-cor.test(correl.df.r2[,i], correl.df.r2[,j], method="spearman")$p.value
    pval.df.r2[i,j] <- x
    }
}
#converting data frame to vector
pval.v.r2<-vector()
for(i in 1:6){
  x<-c(pval.df.r2[i,(i+1):7])
  pval.v.r2<-c(pval.v.r2, x)
}
```

```{r correl.reproducibility.h2o2}
#Are significant correlatons reproducible for two pairs of biological replicates (exponential growth phase vs oxidative stress)?

reprod.v<-vector()
for(i in 1:length(padj.v)){
x<-if(padj.v[i] < 0.05 & (pval.v.r1[i] < 0.05 & pval.v.r2[i] < 0.05)){"TRUE"}else(if(padj.v[i] < 0.05 & (pval.v.r1[i] >= 0.05 | pval.v.r2[i] >= 0.05)){"FALSE"}else{"NA"})
reprod.v<-c(reprod.v, x)
}
reprod.v
```
